### 2015-05-22 Harris_2015/manuscript-materials/draft.Rmd
actually resubmitted

If nontrophic species interactions, such as competition, are important drivers
of community assembly, then ecologists might expect to see their influence in
our data sets [@macarthur_population_1958; @diamond_island_1975]. Despite
decades of work and several major controversies, however [@lewin_santa_1983;
@strong_ecological_1984; @gotelli_swap_2003; @connor_checkered_2013], existing
methods for detecting competition's effects on community structure are
unreliable [@gotelli_empirical_2009], and thus important ecological processes
remain poorly understood. More generally, it can be difficult to reason about
the complex web of direct and indirect species interactions in real
assemblages, especially when these interactions occur against a background of
other ecological processes such as dispersal and environmental filtering
[@connor_checkered_2013]. For this reason, it isn't always clear what kinds of
patterns would even constitute evidence of competition, as opposed to some
other biological process or random sampling error
[@roughgarden_competition_1983; @lewin_santa_1983].

@@

While the competition between the two shrubs in the previous example does not
leave the commonly-expected pattern in community structure (negative
association at the landscape level), it nevertheless does leave a signal in
the data (Figure 1C). Specifically, *among shaded sites*, there will be a
deficit of co-occurrences, and *among unshaded sites*, there will also be such
a deficit.

@@

The rest of the paper proceeds as follows. First, I discuss how Markov networks
work and how they can be used to simulate landscape-level data or to predict
the direct and indirect consequences of possible interaction matrices. Then,
using simulated data sets where the "true" ecological structure is known, I
compare this approach with several existing methods for detecting species
interactions. Finally, I discuss opportunities for extending the approach
presented here to larger problems in community ecology.

@@

Estimating the marginal relationships predicted by a Markov network is more
difficult than estimating conditional relationships, because doing so requires
absolute probability estimates. Turning the relative probability given by
Equation 1 into an absolute probability entails scaling by a *partition
function*, $Z(\alpha, \beta)$, which ensures that the probabilities of all
possible assemblages that could be produced by the model sum to one (bottom of
Figure 2B). Calculating $Z(\alpha, \beta)$ exactly, as is done in this paper,
quickly becomes infeasible as the number of species increases: with
$2^N$ possible assemblages of $N$ species, the number of bookkeeping operations
required for exact inference spirals exponentially into the billions and beyond.
Numerous techniques are available for working with Markov networks that keep
the computations tractable, e.g. via analytic approximations
[@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
but they are beyond the scope of this paper.

@@

***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
The previous sections involved known values of $\alpha$ and $\beta$. In practice,
ecologists will often need to estimate these parameters from data instead. 
When the number of species is reasonably small, one can compute exact maximum 
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing 
$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum. This maximum is the unique combination of $\alpha$ and $\beta$
coefficients that would be expected to produce exactly the observed co-occurrence 
frequencies. For the analyses in this paper, I used the *rosalia* package 
[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to 
define the objective function and gradient as R code.  The rosalia package then 
uses the `BFGS` method in R's `optim` function to find the best values for $\alpha$
and $\beta$.

@@

The first two interaction measures were the sample covariances and the 
partial covariances between each pair of species' data vectors on the landscape
[@albrecht_spatial_2001]. Because partial covariances are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on ridge regression [@van_wieringen_ridge_2014; i.e. linear
regression with a Gaussian prior]. For these analyses, I set the ridge
parameter to 0.2 divided by the number of sites on the landscape.

The third method, described in @gotelli_empirical_2009, involved simulating
possible landscapes from a null model that retains the row and column sums of
the original matrix [@strong_ecological_1984]. Using the default options in the
Pairs software described in @gotelli_empirical_2009, I simulated  the null
distribution of scaled C-scores (a test statistic describing the number of
*non*-co-occurrences between two species). The software then calculated a
$Z$ statistic for each species pair using this null distribution. After
multiplying this statistic by $-1$ so that positive values corresponded to
facilitation and negative values corresponded to competition, I used it as
another estimate of species interactions.

***Method evaluation.***
I evaluated each method qualitatively on the simulated landscapes based on
Figure 1: any method that reliably determined that the two shrub
species were negatively associated passed; other methods failed.

For the larger landscapes, I rescaled the four methods' estimates using linear
regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the "true" $\beta$ coefficient for
each species pair against the model's estimate, re-weighting the pairs so that
each landscape contributed equally to the rescaled estimate. For each  method,
I calculated the squared deviations between each species pair's rescaled
estimate and the "true" interactions used to generate the assemblages.
Finally, I  averaged these squared errors for each combination of species
richness, landscape size, statistical method, and presence/absence of
environmental filtering across all 12 replicates; the mean squared errors
associated with these subsets of the data determined the proportion of
variance explained by each method under different conditions.

@@

***Three species.***
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---
due to indirect effects of the dominant tree species. As a result, the
covariance method falsely reported positive associations on 94% of simulated landscapes,
and the randomization-based null model falsely reported such associations 100%
of the time. The two methods for evaluating conditional relationships (Markov
networks and partial covariances), however, successfully controlled for the
indirect pathway via the tree species and each correctly identified the direct
negative interaction between the shrubs 94% of the time.

***Larger landscapes.***
The accuracy of the four evaluated methods varied substantially, depending
on the parameters that produced the simulated communities (Figure 3).
In general, however, there was a consistent ordering: overall, the Markov network
explained 54% of the variance in the "true" parameters, followed by 
partial covariances (33%), and sample covariances (22%).

The null model scores initially explained only 12% of the variation. After
manually reducing the value of one especially strong outlier ($Z=1004$,
implying $p<10^{-1000000}$), this increased to 17% (Appendix B).  Figure 3
reflects the adjusted version of the results.

@@

The results presented above show that Markov networks, unlike a common null
modeling approach, can recover species' pairwise interactions from
observational data. The Markov networks were successful even when direct
interactions are largely overwhelmed by indirect effects (Figure 1) or
environmental effects (lower panels of Figure 3). For cases where fitting a
Markov network is computationally infeasible, these results also indicate that
partial covariances---which can be computed straightforwardly by linear
regression---can often provide a surprisingly useful approximation. The
success of the partial correlation's success on simulated data may not 
carry over to real data sets, however; @loh_structure_2013 suggest that the linear 
approximations may be less reliable in cases where the true interaction matrix 
contains more structure (e.g. guilds or trophic levels). On the other hand,
if ecologists are familiar enough with the natural history of their study
systems to describe this kind of structure as a prior distribution on the
parameters or as a penalty on the likelihood, then real-world results might
be even better than those shown in Figure 3.

@@

Finally, the results presented here have important implications for ecologists' 
continued use of null models to draw inferences about species interactions. The
small simulated data sets described by Figure 1 show that test statistics
based on marginal co-occurrence (such as C-scores) will not always have a
straightforward relationship with the underlying ecological processes. More
generally, deviations from the null model generally provided less information
about direct species interactions than correlation coefficients did.
Scientists currently using null modeling approaches in their research may be
able to capture twice as much of the variance in species' true interaction
strengths using partial covariances from linear regression instead,  or three
times as much using a Markov network.

Null and neutral models can be very useful for clarifying our thinking about
the numerical consequences of species' richness and abundance patterns
[@harris_occupancy_2011; @xiao_strong_2015], but deviations from a null model
must be interpreted with care [@roughgarden_competition_1983]. In complex
networks of ecological interactions---and even in small networks with three
species---it may simply not be possible to implicate individual species pairs
or specific ecological processes like competition by rejecting a general-
purpose null. Direct estimates of species' conditional associations may be the
only way to make these inferences reliably.

>>>>

If nontrophic species interactions, such as competition, are important drivers
of community assembly, then ecologists might expect to see their influence in
our data sets [@macarthur_population_1958; @diamond_island_1975]. Despite
decades of work and several major controversies, however [@lewin_santa_1983;
@strong_ecological_1984; @gotelli_swap_2003; @connor_checkered_2013], existing
methods for detecting competition's effects on community structure are
unreliable [@gotelli_empirical_2009], and thus important ecological processes
remain poorly understood. More generally, it can be difficult to reason about
the complex web of direct and indirect species interactions in real
assemblages, especially when these interactions occur against a background of
other ecological processes such as dispersal limitation and environmental filtering
[@connor_checkered_2013]. For this reason, it isn't always clear what kinds of
patterns would even constitute evidence of competition, as opposed to some
other biological process or random sampling error
[@roughgarden_competition_1983; @lewin_santa_1983].

@@

While competition between the two shrubs in the previous example does not
leave the commonly-expected pattern in community structure (negative
association at the landscape level), it nevertheless does leave a signal in
the data (Figure 1C). Specifically, *among shaded sites*, there will be a
deficit of co-occurrences, and *among unshaded sites*, there will also be such
a deficit.

@@

The rest of the paper proceeds as follows. First, I discuss how Markov networks
operate and how they can be used to simulate landscape-level data or to predict
the direct and indirect consequences of possible interaction matrices. Then,
using simulated data sets where the "true" ecological structure is known, I
compare this approach with several existing methods for detecting species
interactions. Finally, I discuss opportunities for extending the approach
presented here to larger problems in community ecology.

@@

Estimating the marginal relationships predicted by a Markov network is more
difficult than estimating conditional relationships, because doing so requires
absolute probability estimates. Turning the relative probability given by
Equation 1 into an absolute probability entails scaling by a *partition
function*, $Z(\alpha, \beta)$, which is defined so that that the probabilities of all
possible assemblages that could be produced by the model sum to one (bottom of
Figure 2B). Calculating $Z(\alpha, \beta)$ exactly, as is done in this paper,
quickly becomes infeasible as the number of species increases: with
$2^N$ possible assemblages of $N$ species, the number of bookkeeping operations
required for exact inference spirals exponentially into the billions and beyond.
Numerous techniques are available for working with Markov networks that keep
the computations tractable, e.g. via analytic approximations
[@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
but they are beyond the scope of this paper.

@@

***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
however, ecologists will often need to estimate these parameters from co-occurrence data. 
When the number of species is reasonably small, one can compute exact maximum 
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing 
$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum. This maximum is the unique combination of $\alpha$ and $\beta$
coefficients that would be expected to produce exactly the observed co-occurrence 
frequencies. For the analyses in this paper, I used the *rosalia* package 
[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to 
define the objective function and gradient as R code.  The rosalia package then 
uses the `BFGS` method in R's `optim` function to find the best values for $\alpha$
and $\beta$.

@@

The first two alternative interaction measures were the sample covariances and the 
partial covariances between each pair of species' data vectors on the landscape
[@albrecht_spatial_2001]. Because partial covariances are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on ridge regression [@van_wieringen_ridge_2014; i.e. linear
regression with a Gaussian prior]. For these analyses, I set the ridge
parameter to 0.2 divided by the number of sites on the landscape.

The third alternative method, described in @gotelli_empirical_2009, involved simulating
possible landscapes from a null model that retains the row and column sums of
the original matrix [@strong_ecological_1984]. Using the default options in the
Pairs software described in @gotelli_empirical_2009, I simulated  the null
distribution of scaled C-scores (a test statistic describing the number of
*non*-co-occurrences between two species). The software then calculated a
$Z$ statistic for each species pair using this null distribution. After
multiplying this statistic by $-1$ so that positive values corresponded to
facilitation and negative values corresponded to competition, I used it as
another estimate of species interactions.

***Method evaluation.*** 
For the simulated landscapes based on Figure 1, method evaluation was fairly
qualitative: any method whose test statistic for the two shrubs indicated a
negative relationship passed; other methods failed.

For the larger landscapes, I rescaled the four methods' estimates using linear
regression through the origin so that they all had a consistent
interpretation. In each method, I regressed the "true" $\beta$ coefficient for
each species pair against the model's estimate, re-weighting the pairs so that
each landscape contributed equally to the rescaled estimate. These regressions
yielded squared errors for each of the 23,140 simulated species pairs, across
all conditions and replicates. I then partitioned these errors into groups
defined by the properties of the simulated landscapes (i.e. species richness,
number of observed communities, and the presence/absence of environmental
filtering). Within each partition, the mean squared error was used to
calculate the proportion of variance explained (compared with a baseline model
that assumed all interaction strengths to be zero).

@@

***Three species.*** 
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---due
to indirect effects of the dominant tree species. As a result, the
covariance method falsely reported positive associations for 94% of the simulated
landscapes, and the randomization-based null model falsely reported such
associations 100% of the time. The two methods for evaluating conditional
relationships (Markov networks and partial covariances), however, successfully
controlled for the indirect pathway via the tree species and each correctly
identified the direct negative interaction between the shrubs 94% of the time.

***Larger landscapes.*** 
The accuracy of the four evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3). In general,
however, there was a consistent ordering: overall, the Markov network
explained 54% of the "true" parameters' squared deviations from zero, followed
by partial covariances (33%), and sample covariances (22%). The null model
scores initially explained only 12%. After manually reducing the value of one
especially strong outlier ($Z=1004$, implying $p<10^{-1000000}$), this
increased to 17% (Appendix B).  Figure 3 reflects the adjusted version of the
results.

@@

The results presented above show that Markov networks can reliably recover 
species' pairwise interactions from observational data, even for cases where
a common null modeling technique reliably fails. Specifically, Markov networks 
were successful even when direct interactions were largely overwhelmed by indirect 
effects (Figure 1) or environmental effects (lower panels of Figure 3). For cases 
where fitting a Markov network is computationally infeasible, these results also 
indicate that partial covariances---which can be computed straightforwardly by linear
regression---can often provide a surprisingly useful approximation. The
partial correlations' success on simulated data may not 
carry over to real data sets, however; @loh_structure_2013 show that the linear 
approximations can be less reliable in cases where the true interaction matrix 
contains more structure (e.g. guilds or trophic levels). On the other hand,
if ecologists are familiar enough with the natural history of their study
systems to describe this kind of structure as a prior distribution on the
parameters or as a penalty on the likelihood, then this information could
reduce the effective degres of freedom to estimate and real-world results might
be even better than those shown in Figure 3.

@@

Finally, the results presented here have important implications for
ecologists' continued use of null models to draw inferences about species
interactions. Null and neutral models can be very useful for clarifying our
thinking about the numerical consequences of species' richness and abundance
patterns [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a
null model must be interpreted with care [@roughgarden_competition_1983]. In
complex networks of ecological interactions (and even in small networks with
three species), it may simply not be possible to implicate individual species
pairs or specific ecological processes like competition by rejecting a
general-purpose null [@gotelli_empirical_2009]. Estimating pairwise
coefficients directly seems like a much more promising approach: to the extent
that the models' relative performance on real data sets is similar to the
range of results shown in Figure 3, scientists in this field could easily
double their explanatory power by switching from null models to linear
regression and partial covariances, or triple it by switching to a Markov
network.
